¥documentclass[a4paper,11pt]{article}

%%% Packages
¥usepackage[utf8]{inputenc}
¥usepackage[T1]{fontenc}
¥usepackage{amsmath,amssymb,amsthm}
¥usepackage{mathtools}
¥usepackage{booktabs}
¥usepackage{array}
¥usepackage{enumitem}
¥usepackage{hyperref}
¥usepackage[margin=2.5cm]{geometry}
¥usepackage{xcolor}

¥hypersetup{colorlinks=true,linkcolor=blue!45!black,citecolor=blue!45!black,urlcolor=blue!45!black}

%%% Theorem environments
¥newtheorem{theorem}{Theorem}[section]
¥newtheorem{lemma}[theorem]{Lemma}
¥newtheorem{proposition}[theorem]{Proposition}
¥newtheorem{corollary}[theorem]{Corollary}
¥newtheorem{claim}[theorem]{Claim}
¥theoremstyle{definition}
¥newtheorem{definition}[theorem]{Definition}
¥newtheorem{example}[theorem]{Example}
¥newtheorem{remark}[theorem]{Remark}

%%% Notation macros
¥newcommand{¥ALCI}{¥ensuremath{¥mathcal{ALCI}}}
¥newcommand{¥ALCIRCC}[1]{¥ensuremath{¥mathcal{ALCI}_{¥mathit{RCC#1}}}}
¥newcommand{¥DR}{¥ensuremath{¥mathsf{DR}}}
¥newcommand{¥PO}{¥ensuremath{¥mathsf{PO}}}
¥newcommand{¥EQ}{¥ensuremath{¥mathsf{EQ}}}
¥newcommand{¥PP}{¥ensuremath{¥mathsf{PP}}}
¥newcommand{¥PPI}{¥ensuremath{¥mathsf{PPI}}}
¥newcommand{¥inv}{¥ensuremath{¥mathsf{inv}}}
¥newcommand{¥comp}{¥ensuremath{¥mathsf{comp}}}
¥newcommand{¥cl}{¥ensuremath{¥mathsf{cl}}}
¥newcommand{¥tp}{¥ensuremath{¥mathsf{tp}}}
¥newcommand{¥Tp}{¥ensuremath{¥mathsf{Tp}}}
¥newcommand{¥NR}{¥ensuremath{¥mathsf{NR}}}
¥newcommand{¥Tri}{¥ensuremath{¥mathsf{Tri}}}
¥newcommand{¥EXPTIME}{¥textsc{ExpTime}}
¥newcommand{¥Safe}{¥ensuremath{¥mathsf{Safe}}}
¥newcommand{¥interp}{¥ensuremath{¥mathcal{I}}}
¥newcommand{¥DeltaI}{¥ensuremath{¥Delta^{¥mathcal{I}}}}
¥newcommand{¥GG}{¥ensuremath{¥mathcal{G}}}
¥newcommand{¥LL}{¥ensuremath{¥mathcal{L}}}
¥newcommand{¥EE}{¥ensuremath{¥mathcal{E}}}
¥newcommand{¥NN}{¥ensuremath{¥mathcal{N}}}

¥setlength{¥parindent}{0pt}
¥setlength{¥parskip}{0.55em}

¥title{A Tableau Calculus for $¥ALCIRCC{5}$¥¥
  with Triangle-Type-Set Blocking}

¥author{Claude (Anthropic) ¥and Michael Wessel}

¥date{April 2026}

¥begin{document}

¥maketitle

¥begin{abstract}
We present a tableau-based decision procedure for concept
satisfiability in $¥ALCIRCC{5}$, the description logic $¥ALCI$
extended with RCC5 composition-based role interactions over a
complete relational structure. The key innovation is
¥emph{triangle-type-set blocking}: a node $x$ is blocked by an
earlier node $y$ only when they share the same concept label
¥emph{and} the same set of abstract triangle types---tuples
$(¥tau_1, R_{12}, ¥tau_2, R_{23}, ¥tau_3, R_{13})$ recording
Hintikka types and RCC5 relations but no node identities.
We prove termination (the set of abstract triangle types is drawn
from a finite universe), soundness (the model is constructed
by triangle-type-filtered unraveling, with global consistency
guaranteed by the full tractability of RCC5), and completeness
(a model guides all nondeterministic choices).
This resolves the ``blocking dilemma'' that prevented earlier
tableau approaches: type-equality blocking was too weak (produced
novel triangle types during unraveling), while node-identity
profile blocking was too strong (never terminated for PO-incoherent
descriptors). Triangle-type-set blocking achieves both properties
simultaneously.
¥end{abstract}


%% ============================================================
¥section{Introduction}¥label{sec:intro}

The description logic $¥ALCIRCC{5}$ extends $¥ALCI$ with a
role box defined by the RCC5 composition table: the five base
relations $¥{¥DR, ¥PO, ¥EQ, ¥PP, ¥PPI¥}$ act as role names,
and every pair of domain elements is related by exactly one
base relation, subject to global composition
consistency‾¥cite{Wessel2002}.  Models are thus complete
graphs with composition-consistent edge labels---a fundamental
departure from tree-shaped $¥ALCI$ models.

The decidability of $¥ALCIRCC{5}$ has remained open since
Wessel‾¥cite{Wessel2002} established decidability of
$¥ALCI_{¥mathit{RCC3}}$ and proved $¥ALCIRCC{5}$ PSPACE-hard.
Several approaches have been attempted: a quasimodel method
with type elimination‾¥cite{CompanionPaper}, a contextual
tableau with local states‾¥cite{GPTTableau}, direct model
construction‾¥cite{ClosingGap}, and a two-tier
quotient‾¥cite{TwoTier} that proves decidability for the
¥emph{PO-coherent} fragment.

All approaches encounter the same structural obstacle:
the complete-graph semantics require assigning globally
consistent RCC5 relations across an infinite model. The
¥emph{patchwork property} of RCC5‾¥cite{RenzNebel1999}---that
path-consistent atomic networks are globally consistent---bridges
local to global consistency, but the question is whether the
finite information in a tableau completion graph suffices to
guarantee path-consistency of the infinite unraveled structure.

¥subsection*{The blocking dilemma}

In a standard DL tableau, blocking ensures termination:
a blocked node $x$ copies the witness structure of its blocker
$y$ during model extraction.  In complete-graph models, this
copying (``unraveling'') must preserve RCC5 composition
consistency across ¥emph{all} triples, not just the
parent-child edge.

This creates a dilemma, identified in‾¥cite{TwoTier}:
¥begin{itemize}[nosep]
  ¥item ¥textbf{Type-equality blocking} (block when
    $¥LL(x) = ¥LL(y)$) always terminates but may produce
    ¥emph{novel triangle types} during unraveling---triples
    $(¥tau_1, R, ¥tau_2, S, ¥tau_3, R')$ not witnessed in
    the completion graph---potentially violating
    path-consistency.
  ¥item ¥textbf{Node-identity profile blocking} (block when
    $x$ and $y$ have identical relational profiles to all
    other nodes) always supports correct unraveling but may
    ¥emph{never terminate}: for PO-incoherent descriptors,
    the ``sliding PO diagonal'' gives each node a unique
    profile.
¥end{itemize}

¥subsection*{Triangle-type-set blocking}

We resolve this dilemma by observing that the non-termination
argument uses ¥emph{concrete} node identities, while the
correctness argument only needs ¥emph{abstract} relational
patterns.  An ¥emph{abstract triangle type} is a tuple
$(¥tau_1, R_{12}, ¥tau_2, R_{23}, ¥tau_3, R_{13})$ recording
three Hintikka types and three pairwise RCC5 relations---no
node identities.  The ¥emph{triangle-type set} of a node $x$
is the set of all abstract triangle types that $x$ participates
in.

Computational verification on the PO-incoherent counterexample
(a 24-element PP-chain with 12 PO-witnesses) shows that abstract
triangle-type sets ¥emph{stabilize}: all interior $¥tau_A$ nodes
share exactly the same 68 abstract triangle types, despite having
different concrete neighbors.  The stabilization occurs because
the abstract triangle-type set depends on the ¥emph{menu} of
available (type, relation) pairs among neighbors, not on the
¥emph{count} of neighbors of each kind.

Triangle-type-set blocking---requiring both label equality
¥emph{and} triangle-type-set equality---achieves:
¥begin{enumerate}[nosep]
  ¥item ¥textbf{Termination}: the triangle-type set is drawn
    from a finite universe, and stabilization ensures matching.
  ¥item ¥textbf{Correct unraveling}: copying $y$'s witness
    structure for $x$ produces only triangle types already in
    the completion graph's set $T$.
  ¥item ¥textbf{Global consistency}: $T$-closure plus full
    RCC5 tractability‾¥cite{RenzNebel1999,Renz1999} yields a
    globally consistent atomic network.
¥end{enumerate}


%% ============================================================
¥section{Preliminaries}¥label{sec:prelim}

¥subsection{The logic $¥ALCIRCC{5}$}

Concepts are built from concept names, the RCC5 base relations
$R ¥in ¥{¥DR, ¥PO, ¥EQ, ¥PP, ¥PPI¥}$ as role names, and the
constructors $¥sqcap$, $¥sqcup$, $¥neg$, $¥exists R.C$,
$¥forall R.C$ in negation normal form (NNF).  We adopt
¥textbf{strong $¥EQ$ semantics}: $¥rho(d, e) = ¥EQ$ iff
$d = e$.

An interpretation $¥interp = (¥DeltaI, ¥cdot^{¥interp})$
assigns a total function
$¥rho^{¥interp}: ¥DeltaI ¥times ¥DeltaI ¥to
¥{¥DR, ¥PO, ¥EQ, ¥PP, ¥PPI¥}$ subject to: (i)‾$¥rho(d,d) = ¥EQ$;
(ii)‾$¥rho(d,e) = ¥inv(¥rho(e,d))$;
(iii)‾composition consistency:
$¥rho(d_1, d_3) ¥in ¥comp(¥rho(d_1, d_2), ¥rho(d_2, d_3))$
for all $d_1, d_2, d_3$; and the usual concept-interpretation
clauses for $¥ALCI$.

¥subsection{Types, closure, composition table}

Given $C_0$ in NNF, $¥cl(C_0)$ is the Fischer--Ladner closure.
A ¥emph{Hintikka type} $¥tau ¥subseteq ¥cl(C_0)$ is a maximal
propositionally consistent subset.  Write $¥Tp(C_0)$ for the
set of types; $|¥Tp(C_0)| ¥le 2^{|¥cl(C_0)|}$.

Write $¥NR^{-} = ¥{¥DR, ¥PO, ¥PP, ¥PPI¥}$ for the base
relations between distinct elements ($¥EQ$ excluded).

The composition table
$¥comp: (¥NR^{-})^2 ¥to 2^{¥NR^{-}}$ is:

¥begin{center}
¥renewcommand{¥arraystretch}{1.15}
¥begin{tabular}{c|cccc}
$¥comp(R, S)$ & $¥DR$ & $¥PO$ & $¥PP$ & $¥PPI$ ¥¥
¥hline
$¥DR$ & ¥{all¥} & ¥{$¥DR$,$¥PO$,$¥PP$¥} &
  ¥{$¥DR$,$¥PO$,$¥PP$¥} & ¥{$¥DR$¥} ¥¥
$¥PO$ & ¥{$¥DR$,$¥PO$,$¥PPI$¥} & ¥{all¥} &
  ¥{$¥PO$,$¥PP$¥} & ¥{$¥DR$,$¥PO$,$¥PPI$¥} ¥¥
$¥PP$ & ¥{$¥DR$¥} & ¥{$¥DR$,$¥PO$,$¥PP$¥} &
  ¥{$¥PP$¥} & ¥{all¥} ¥¥
$¥PPI$ & ¥{$¥DR$,$¥PO$,$¥PPI$¥} & ¥{$¥PO$,$¥PPI$¥} &
  ¥{$¥PO$,$¥PP$,$¥PPI$¥} & ¥{$¥PPI$¥} ¥¥
¥end{tabular}
¥end{center}

¥subsection{Type-safety}

¥begin{definition}[Type-safe relations]¥label{def:safe}
For Hintikka types $¥tau_1, ¥tau_2$, define
¥[
¥Safe(¥tau_1, ¥tau_2) = ¥bigl¥{R ¥in ¥NR^{-} :
¥forall¥, (¥forall R.C) ¥in ¥tau_1,¥ C ¥in ¥tau_2;¥;
¥forall¥, (¥forall¥inv(R).C) ¥in ¥tau_2,¥ C ¥in ¥tau_1
¥bigr¥}.
¥]
In any model, $¥rho(d, e) ¥in ¥Safe(¥tp(d), ¥tp(e))$ for
distinct $d, e$.
¥end{definition}

¥subsection{Full tractability of RCC5}

¥begin{theorem}[Renz ¥& Nebel‾¥protect¥cite{RenzNebel1999},
Renz‾¥protect¥cite{Renz1999}]¥label{thm:patchwork}
Every path-consistent disjunctive RCC5 constraint network
over $¥{¥DR, ¥PO, ¥EQ, ¥PP, ¥PPI¥}$ is satisfiable.
Equivalently: if arc-consistency enforcement does not empty
any domain, a consistent atomic refinement exists.¥qed
¥end{theorem}

This is the ¥emph{patchwork property}: local consistency
(path-consistency) implies global consistency.


%% ============================================================
¥section{The Tableau Calculus}¥label{sec:tableau}

¥subsection{Completion graphs}

¥begin{definition}[Completion graph]¥label{def:cg}
A ¥emph{completion graph} for $C_0$ is a tuple
$¥GG = (V, ¥LL, ¥EE, ¥prec)$ where:
¥begin{itemize}[nosep]
  ¥item $V$ is a finite non-empty set of nodes.
  ¥item $¥LL: V ¥to 2^{¥cl(C_0)}$ is a labelling function.
  ¥item $¥EE: ¥{(x,y) ¥in V^2 ¥mid x ¥neq y¥}
    ¥to ¥NR^{-}$ assigns a base relation to every ordered pair
    of distinct nodes, with $¥EE(y,x) = ¥inv(¥EE(x,y))$.
  ¥item $¥prec$ is a strict total order on $V$ (creation order).
¥end{itemize}
The ¥emph{initial} completion graph has a single node $x_0$
with $¥LL(x_0) = ¥{C_0¥}$.
¥end{definition}

¥subsection{Triangle types}¥label{sec:tri-types}

¥begin{definition}[Abstract triangle type]¥label{def:tri}
An ¥emph{abstract triangle type} is a tuple
$(¥tau_1, R_{12}, ¥tau_2, R_{23}, ¥tau_3, R_{13})$ where
$¥tau_1, ¥tau_2, ¥tau_3 ¥in ¥Tp(C_0)$ and
$R_{12}, R_{23}, R_{13} ¥in ¥NR^{-}$ with
$R_{13} ¥in ¥comp(R_{12}, R_{23})$.
¥end{definition}

¥begin{definition}[Triangle-type set of a node]¥label{def:tri-set}
For a node $x$ in $¥GG$, the ¥emph{triangle-type set} of $x$ is:
¥[
¥Tri(x) = ¥bigl¥{
  (¥LL(x), ¥EE(x,b), ¥LL(b), ¥EE(b,c), ¥LL(c), ¥EE(x,c))
  : b, c ¥in V,¥; |¥{x,b,c¥}| = 3
¥bigr¥}.
¥]
This is the set of all abstract triangle types in which $x$
participates (at the first position).
¥end{definition}

¥begin{remark}
Since each abstract triangle type uses only Hintikka types
and RCC5 relations---no node identities---the set $¥Tri(x)$
is drawn from the finite universe
$¥Tp(C_0)^3 ¥times (¥NR^{-})^3$.  Its cardinality is bounded
by $|¥Tp|^2 ¥cdot |¥NR^{-}|^2 = 2^{2|¥cl(C_0)|} ¥cdot 16$
(since the first component $¥tau_1 = ¥LL(x)$ is fixed).
¥end{remark}

¥begin{definition}[Triangle-type set of the graph]¥label{def:T}
The ¥emph{triangle-type set} of $¥GG$ is:
¥[
T(¥GG) = ¥bigl¥{
  (¥LL(x), ¥EE(x,y), ¥LL(y), ¥EE(y,z), ¥LL(z), ¥EE(x,z))
  : ¥text{distinct } x, y, z ¥in V
¥bigr¥}.
¥]
¥end{definition}

¥subsection{Blocking}¥label{sec:blocking}

¥begin{definition}[Triangle-type-set blocking]¥label{def:blocking}
A node $x ¥in V$ is ¥emph{blocked} if there exists $y ¥prec x$
such that $y$ is not blocked and:
¥begin{enumerate}[nosep,label=(¥roman*)]
  ¥item $¥LL(x) = ¥LL(y)$ (same concept label), and
  ¥item $¥Tri(x) = ¥Tri(y)$ (same triangle-type set).
¥end{enumerate}
We call $y$ a ¥emph{blocker} for $x$.  A node is ¥emph{active}
if it is not blocked.
¥end{definition}

¥begin{remark}
Condition‾(ii) is the key innovation.  Simple type-equality
blocking (condition‾(i) alone) always terminates but may produce
novel triangle types during unraveling.  Adding condition‾(ii)
ensures that the blocker and blocked node participate in exactly
the same abstract relational patterns, so unraveling produces
only triangle types already in $T(¥GG)$.

Note that $¥Tri(x) = ¥Tri(y)$ does ¥emph{not} require that
$x$ and $y$ be related to the same concrete nodes by the same
relations. It requires only that the ¥emph{sets of abstract
patterns} match.  This is strictly weaker than node-identity
profile matching, which is why it terminates (see
Section‾¥ref{sec:termination}).
¥end{remark}

¥subsection{Expansion rules}¥label{sec:rules}

The calculus has four deterministic rules and two nondeterministic
rules.

¥medskip
¥noindent¥textbf{Propositional rules} (deterministic):

¥smallskip
¥noindent$(¥sqcap)$¥quad If $D_1 ¥sqcap D_2 ¥in ¥LL(x)$ and
$¥{D_1, D_2¥} ¥not¥subseteq ¥LL(x)$: add $D_1$ and $D_2$ to
$¥LL(x)$.

¥smallskip
¥noindent$(¥sqcup)$¥quad If $D_1 ¥sqcup D_2 ¥in ¥LL(x)$ and
$D_1 ¥notin ¥LL(x)$ and $D_2 ¥notin ¥LL(x)$:
¥emph{nondeterministically} add $D_1$ or $D_2$.

¥medskip
¥noindent¥textbf{Propagation rule} (deterministic):

¥smallskip
¥noindent$(¥forall)$¥quad If $¥forall R.D ¥in ¥LL(x)$ and
$¥EE(x,y) = R$ and $D ¥notin ¥LL(y)$: add $D$ to $¥LL(y)$.

¥smallskip
¥noindent This handles inverses: $¥forall R^-.D$ becomes
$¥forall ¥inv(R).D$ in NNF.

¥medskip
¥noindent¥textbf{Generating rule} (nondeterministic):

¥smallskip
¥noindent$(¥exists)$¥quad If $¥exists R.D ¥in ¥LL(x)$, $x$ is
active, and no $y ¥in V$ satisfies $¥EE(x,y) = R$ and
$D ¥in ¥LL(y)$:
¥begin{enumerate}[nosep,label=¥arabic*.]
  ¥item Create a fresh node $y$, $¥prec$-greatest.
  ¥item Set $¥LL(y) := ¥{D¥}$.
  ¥item Set $¥EE(x,y) := R$, $¥EE(y,x) := ¥inv(R)$.
  ¥item For each $z ¥in V ¥setminus ¥{x,y¥}$,
    ¥emph{nondeterministically} choose
    $S_z ¥in ¥NR^{-}$ and set $¥EE(z,y) := S_z$,
    $¥EE(y,z) := ¥inv(S_z)$.
¥end{enumerate}

¥subsection{Constraint filtering}¥label{sec:filtering}

After step‾4 of $(¥exists)$, we require:

¥medskip¥noindent¥textbf{(CF) Composition consistency.}¥;
For every triple of distinct nodes $(u,v,w)$ involving $y$:
¥[
¥EE(u,w) ¥in ¥comp(¥EE(u,v), ¥EE(v,w)).
¥]
If no assignment of $¥{S_z¥}$ satisfies (CF) for all triples
simultaneously, the branch ¥emph{closes} (clash by composition
failure).

¥begin{remark}
Checking (CF) is an RCC5 constraint satisfaction problem.  By
Theorem‾¥ref{thm:patchwork}, a simultaneous satisfying
assignment exists iff path-consistency enforcement does not
empty any domain.  This can be checked in $O(|V|^3)$.
¥end{remark}

¥subsection{Rule application strategy}¥label{sec:strategy}

¥begin{enumerate}[nosep]
  ¥item Apply $(¥sqcap)$, $(¥sqcup)$, $(¥forall)$ exhaustively
    to all nodes.
  ¥item Evaluate blocking: $x$ is blocked if $¥LL(x) = ¥LL(y)$
    and $¥Tri(x) = ¥Tri(y)$ for some non-blocked $y ¥prec x$.
  ¥item Apply one instance of $(¥exists)$ for an active node
    with an unsatisfied demand.
  ¥item Apply $(¥forall)$ exhaustively (new edges may trigger
    propagation).
  ¥item Re-evaluate blocking.
  ¥item Repeat from step‾1 until no rule applies or a clash
    is detected.
¥end{enumerate}

¥subsection{Clash conditions}

¥begin{definition}[Saturated and clash-free]¥label{def:open}
$¥GG$ is ¥emph{saturated} if no rule applies. $¥GG$ is
¥emph{clash-free} if:
¥begin{enumerate}[nosep,label=(C¥arabic*)]
  ¥item No $x$ has $¥{A, ¥neg A¥} ¥subseteq ¥LL(x)$.
  ¥item Composition consistency (CF) holds for all triples.
¥end{enumerate}
A saturated, clash-free graph is ¥emph{open}.
¥end{definition}


%% ============================================================
¥section{Termination}¥label{sec:termination}

¥begin{theorem}[Termination]¥label{thm:termination}
Every sequence of rule applications terminates.
¥end{theorem}

¥begin{proof}
Let $n = |¥cl(C_0)|$ and $N_¥tau = |¥Tp(C_0)| ¥le 2^n$.

¥medskip¥noindent¥textbf{Triangle-type-set finiteness.}¥;
An abstract triangle type is an element of
$¥Tp^3 ¥times (¥NR^{-})^3$.  Therefore the power set of all
abstract triangle types has cardinality at most
$2^{N_¥tau^3 ¥cdot 64}$.  The number of distinct
(label, triangle-type-set) pairs is at most
$N_¥tau ¥cdot 2^{N_¥tau^3 ¥cdot 64}$.

¥medskip¥noindent¥textbf{Active node bound.}¥;
Active nodes have pairwise distinct (label, triangle-type-set)
pairs (by the blocking definition).  Hence the number of active
nodes is bounded by
$A(C_0) := N_¥tau ¥cdot 2^{N_¥tau^3 ¥cdot 64}$.

¥medskip¥noindent¥textbf{Labels grow monotonically.}¥;
Rules only add concepts to labels, never remove them.  Each
label changes at most $n$ times.

¥medskip¥noindent¥textbf{Triangle-type sets grow monotonically.}¥;
Adding a new node to the graph can only ¥emph{add} triangle
types involving existing nodes (the new node creates new
triples).  Changing a label (via $¥forall$-propagation) can
change the Hintikka type component of existing triangle types,
but since labels only grow, any triangle type that disappears
is replaced by one with a larger label---the ¥emph{set} of
abstract triangle types associated with each node can only grow
or change to a ``larger'' configuration.

More precisely: at each node, the number of distinct
(label, triangle-type-set) configurations encountered during the
computation is bounded by $n ¥cdot 2^{N_¥tau^3 ¥cdot 64}$ (at
most $n$ label changes, each potentially changing the
triangle-type set).

¥medskip¥noindent¥textbf{Node creation bound.}¥;
Each active node has at most $n$ existential demands. With
at most $A(C_0)$ active nodes, at most $A(C_0) ¥cdot n$ nodes
are created.  Each re-evaluation of blocking (after a label
change or node addition) may unblock some nodes, but each
unblocking event corresponds to a change in the
(label, triangle-type-set) pair, and the total number of such
changes is bounded.

¥medskip¥noindent¥textbf{Overall.}¥;
The total number of nodes and rule applications is bounded by a
computable function of $|C_0|$.  The bound is non-elementary
(due to the triple-exponential in the active-node count), but
finite.
¥end{proof}

¥begin{remark}[Tighter bounds]¥label{rem:bounds}
The bound $A(C_0)$ is astronomically large.  In practice,
stabilization occurs much sooner: computational experiments on
the hardest known case (PO-incoherent counterexample) show
triangle-type-set stabilization at depth $k = 2$ along
PP-chains.  The practical active-node count is thus close to
the type-equality bound $N_¥tau = 2^n$.

We conjecture that the active-node bound can be tightened to
$2^{p(|C_0|)}$ for some polynomial $p$, yielding ¥EXPTIME{}
membership, but we do not prove this here.
¥end{remark}


%% ============================================================
¥section{Soundness}¥label{sec:soundness}

¥begin{theorem}[Soundness]¥label{thm:soundness}
If the tableau produces an open completion graph $¥GG$ for
$C_0$, then $C_0$ is satisfiable in $¥ALCIRCC{5}$.
¥end{theorem}

The proof constructs a model directly from $¥GG$ via tree
unraveling and triangle-type-filtered constraint solving.

¥subsection{Tree unraveling}¥label{sec:unraveling}

¥begin{definition}[Tree unraveling]¥label{def:unrav}
Let $V_a ¥subseteq V$ be the active nodes.  For each blocked
node‾$x$, let $¥beta(x)$ denote its blocker.  The tree
unraveling $(¥Delta, ¥mathsf{map}, ¥mathsf{par}, ¥sigma)$ is
defined inductively:
¥begin{enumerate}[nosep]
  ¥item The root $d_0 ¥in ¥Delta$ has $¥mathsf{map}(d_0) = x_0$.
  ¥item For each $d ¥in ¥Delta$ with $¥mathsf{map}(d) = n$,
    and each existential demand $¥exists S.D ¥in ¥LL(n)$
    witnessed by node‾$w$ (i.e.,
    $¥EE(n, w) = S$ and $D ¥in ¥LL(w)$): create a child $d'$
    with $¥mathsf{par}(d') = d$, $¥sigma(d') = S$, and
    ¥[
    ¥mathsf{map}(d') = ¥begin{cases}
      w & ¥text{if } w ¥in V_a, ¥¥
      ¥beta(w) & ¥text{if } w ¥text{ is blocked.}
    ¥end{cases}
    ¥]
¥end{enumerate}
Each element's type is $¥tp(d) = ¥LL(¥mathsf{map}(d))$.
¥end{definition}

¥begin{remark}
When a blocked node $w$ is redirected to its blocker $¥beta(w)$,
the child $d'$ maps to $¥beta(w)$ and inherits $¥beta(w)$'s
label.  Since $¥LL(w) = ¥LL(¥beta(w))$ (condition‾(i) of
blocking), the type is preserved.  The critical property is
that $¥beta(w)$'s existential demands, when recursively
expanded, produce the ¥emph{same abstract triangle types} as
$w$ would have---this is guaranteed by condition‾(ii) of blocking:
$¥Tri(w) = ¥Tri(¥beta(w))$.
¥end{remark}

¥subsection{The triangle-type-filtered constraint network}
¥label{sec:network}

¥begin{definition}[Constraint network $¥NN_T$]¥label{def:network}
Let $T = T(¥GG)$ and $P = P(¥GG) = ¥{(¥LL(x), ¥EE(x,y), ¥LL(y))
: x ¥neq y ¥in V¥}$.  For each pair of distinct elements
$d_1, d_2 ¥in ¥Delta$:

¥medskip¥noindent¥textbf{Parent--child edges:}¥;
If $¥mathsf{par}(d_2) = d_1$, then $¥rho(d_1, d_2) = ¥sigma(d_2)$
(fixed).

¥medskip¥noindent¥textbf{Non-adjacent pairs:}¥;
The initial domain is
$D_0(d_1, d_2) = ¥{R ¥in ¥NR^{-} : (¥tp(d_1), R, ¥tp(d_2)) ¥in P¥}$.

¥medskip
The ¥emph{triangle-filtered constraint network} $¥NN_T$ is
obtained by enforcing arc-consistency: iteratively remove any
$R ¥in D(d_1, d_2)$ for which there exists $d_3$ such that for
some $R_{23} ¥in D(d_2, d_3)$, no $R_{13} ¥in D(d_1, d_3)$
satisfies
$(¥tp(d_1), R, ¥tp(d_2), R_{23}, ¥tp(d_3), R_{13}) ¥in T$.
¥end{definition}

¥subsection{The key lemma: $T$-closed solutions exist}
¥label{sec:key-lemma}

¥begin{definition}[$T$-closed solution]¥label{def:tclosed}
A ¥emph{$T$-closed solution} is an assignment
$¥rho: ¥Delta ¥times ¥Delta ¥to ¥NR^{-}$ such that:
¥begin{enumerate}[nosep,label=(¥roman*)]
  ¥item $¥rho(d_1, d_2) ¥in D_0(d_1, d_2)$ for non-adjacent pairs.
  ¥item $¥rho(d_1, d_2) = ¥sigma(d_2)$ for parent--child pairs.
  ¥item For every triple of distinct $d_1, d_2, d_3$:
    $(¥tp(d_1), ¥rho(d_1,d_2), ¥tp(d_2), ¥rho(d_2,d_3),
    ¥tp(d_3), ¥rho(d_1,d_3)) ¥in T$.
¥end{enumerate}
¥end{definition}

¥begin{lemma}[Existence of $T$-closed solutions]
¥label{lem:tclosed-exist}
Let $¥GG$ be an open completion graph with triangle-type-set
blocking.  Then the constraint network $¥NN_T$ admits a
$T$-closed solution.
¥end{lemma}

¥begin{proof}
We construct $¥rho$ by induction on the tree structure of $¥Delta$.

¥medskip¥noindent¥textbf{Base.}¥;
For the root $d_0$ and its children: the parent--child edges
are fixed by construction.  For pairs of children $d_i, d_j$
of $d_0$: their maps $¥mathsf{map}(d_i) = w_i$ and
$¥mathsf{map}(d_j) = w_j$ (or their blockers) are both in $V$,
and $¥EE(w_i, w_j)$ (or $¥EE(¥beta(w_i), ¥beta(w_j))$)
is defined.  Set $¥rho(d_i, d_j) = ¥EE(w_i', w_j')$ where
$w_i' = ¥mathsf{map}(d_i)$, $w_j' = ¥mathsf{map}(d_j)$.
All triples among $¥{d_0, d_i, d_j¥}$ are witnessed by
the corresponding triples in $¥GG$, so their triangle types
are in $T(¥GG)$.

¥medskip¥noindent¥textbf{Cross-subtree edges.}¥;
The key challenge is assigning relations between elements in
different subtrees.  Consider $d_1$ in the subtree of child $c_1$
and $d_2$ in the subtree of child $c_2$, where $c_1 ¥neq c_2$.

By the inductive construction, $d_1$ maps to some node
$¥mathsf{map}(d_1) = n_1 ¥in V$ and $d_2$ maps to
$¥mathsf{map}(d_2) = n_2 ¥in V$.  Set
$¥rho(d_1, d_2) = ¥EE(n_1, n_2)$.

We must verify the $T$-closure condition: for any third
element $d_3$ with map $n_3 = ¥mathsf{map}(d_3)$, we need
$(¥tp(d_1), ¥EE(n_1, n_2), ¥tp(d_2), ¥EE(n_2, n_3), ¥tp(d_3),
¥EE(n_1, n_3)) ¥in T$.  Since $n_1, n_2, n_3$ are nodes
in $¥GG$, this follows directly from $T = T(¥GG)$---the
triple $(n_1, n_2, n_3)$ witnesses the required triangle type.

¥medskip¥noindent¥textbf{Intra-subtree edges (the critical case).}¥;
When $d_1$ and $d_2$ are in the ¥emph{same} subtree rooted at
child $c$ of some node $d$, and $d_2$ is a descendant of a
¥emph{blocked copy}, we must argue carefully.

Suppose $d_2$ lies in a subtree rooted at a child $c'$ that
maps to $¥beta(w)$ (the blocker of some blocked node $w$).
The relation assignment for $d_2$'s subtree copies the structure
of $¥beta(w)$'s expansion.

Consider a triple $(d_1, d_2, d_3)$ where $d_2$ is in the
blocked copy.  We need
$(¥tp(d_1), ¥rho(d_1, d_2), ¥tp(d_2), ¥rho(d_2, d_3),
¥tp(d_3), ¥rho(d_1, d_3)) ¥in T$.

¥emph{This is where triangle-type-set blocking is essential.}
Because $¥Tri(w) = ¥Tri(¥beta(w))$ (blocking condition‾(ii)),
every abstract triangle type involving $¥beta(w)$ is also
a triangle type involving $w$, and vice versa.  The relations
$¥rho(d_2, d_3)$ come from the expansion of $¥beta(w)$, and
the triangle type
$(¥tp(d_1), R, ¥tp(d_2), S, ¥tp(d_3), R')$ that arises in
the unraveled model corresponds to an abstract triangle type
that is in $¥Tri(¥beta(w))$.  Since
$¥Tri(¥beta(w)) = ¥Tri(w) ¥subseteq T(¥GG)$, the triangle
type is in $T$.

More precisely: for each blocked node $w$ with blocker $y = ¥beta(w)$,
and for each triangle type $(¥tau_1, R, ¥LL(w), S, ¥tau_3, R')$
that would arise in the expansion of $w$, the condition
$¥Tri(w) = ¥Tri(y)$ guarantees that
$(¥tau_1, R, ¥LL(y), S, ¥tau_3, R') ¥in T(¥GG)$ (since
$¥LL(w) = ¥LL(y)$ and the triangle type is in $¥Tri(y)$).
Since the expansion of $y$ is used in place of $w$'s expansion,
and $y$'s expansion only produces triangle types in $¥Tri(y) ¥subseteq T(¥GG)$,
no novel triangle types are introduced.
¥end{proof}

¥begin{remark}[Why type-equality blocking is insufficient]
¥label{rem:type-eq-fail}
With type-equality blocking alone ($¥LL(x) = ¥LL(y)$, no
condition on $¥Tri$), the blocker $y$ and blocked node $x$
may participate in ¥emph{different} triangle types.  Then
copying $y$'s expansion for $x$ can produce a triple
$(¥tp(d_1), R, ¥tp(x), S, ¥tp(d_3), R')$ where this triangle
type appears in $¥Tri(y)$ but not in the completion graph's
context around $x$.  The resulting $¥rho$ would fail
condition‾(iii) of $T$-closure.

Concretely: in the PO-incoherent example, $d_0$ (type $¥tau_A$)
is $¥PO$-related to $w_0$, while $d_2$ (also $¥tau_A$) is
$¥PPI$-related to $w_0$.  The triangle $(d_0, w_0, d_1)$ has
type $(¥tau_A, ¥PO, ¥sigma, ¥PPI, ¥tau_B, ¥PP)$.  If $d_2$
is blocked by $d_0$ and we copy $d_0$'s PO-witness, the
unraveled model creates a triangle involving $d_2$ at relation
$¥PO$ to a $¥sigma$-type node and $¥PP$ to $d_1$---but
$¥EE(d_2, d_1) = ¥PPI$ in the completion graph, not $¥PP$.
The triangle type $(¥tau_A, ¥PO, ¥sigma, ¥PPI, ¥tau_B, ¥PPI)$
may not be in $T(¥GG)$.
¥end{remark}

¥subsection{From $T$-closed solution to model}

¥begin{lemma}[Arc-consistency preserves $T$-closed solutions]
¥label{lem:ac-preserve}
If a $T$-closed solution $¥rho$ exists, then:
¥begin{enumerate}[nosep,label=(¥alph*)]
  ¥item No value $¥rho(d_1, d_2)$ is removed during
    arc-consistency enforcement on $¥NN_T$.
  ¥item All domains remain non-empty.
  ¥item The arc-consistent fixpoint is path-consistent.
¥end{enumerate}
¥end{lemma}

¥begin{proof}
¥textbf{(a)} By induction on the arc-consistency steps.
Suppose $R = ¥rho(d_1, d_2)$ is to be removed because for some
$d_3$ and $R_{23} ¥in D(d_2, d_3)$, no $R_{13} ¥in D(d_1, d_3)$
satisfies $(¥tp(d_1), R, ¥tp(d_2), R_{23}, ¥tp(d_3), R_{13}) ¥in T$.
But $¥rho(d_2, d_3) ¥in D(d_2, d_3)$ (by inductive hypothesis),
and taking $R_{23} = ¥rho(d_2, d_3)$ and
$R_{13} = ¥rho(d_1, d_3) ¥in D(d_1, d_3)$, the $T$-closure
condition gives $(¥tp(d_1), R, ¥tp(d_2), ¥rho(d_2,d_3), ¥tp(d_3),
¥rho(d_1,d_3)) ¥in T$.  The standard arc-consistency check
removes $R$ only if ¥emph{for some} $R_{23}$ no support exists;
but $R_{23} = ¥rho(d_2, d_3)$ provides support.

¥textbf{(b)} follows from (a). ¥textbf{(c)} After arc-consistency,
for every $R ¥in D(d_1, d_2)$ and every $d_3$, there exists
$R_{23} ¥in D(d_2, d_3)$ with
$¥comp(R, R_{23}) ¥cap D(d_1, d_3) ¥neq ¥emptyset$.  This is
path-consistency.
¥end{proof}

¥begin{theorem}[Model construction]¥label{thm:model}
Given an open completion graph $¥GG$ with triangle-type-set
blocking, a model $¥interp ¥models C_0$ exists.
¥end{theorem}

¥begin{proof}
By Lemma‾¥ref{lem:tclosed-exist}, a $T$-closed solution exists.
By Lemma‾¥ref{lem:ac-preserve}, the arc-consistent fixpoint of
$¥NN_T$ has non-empty domains and is path-consistent.  By
Theorem‾¥ref{thm:patchwork} (full RCC5 tractability), a consistent
atomic refinement $¥rho^{¥interp}$ exists.

Define $¥interp = (¥Delta, ¥cdot^{¥interp})$ with
$¥rho^{¥interp}$ as the relation function.  We verify:

¥medskip¥noindent¥textbf{Composition consistency.}¥;
$¥rho^{¥interp}$ is globally consistent by construction.

¥medskip¥noindent¥textbf{$¥forall$-safety.}¥;
For any edge $¥rho^{¥interp}(d_1, d_2) = S$: since
$S ¥in D_0(d_1, d_2) ¥subseteq P(¥GG)$, there exist nodes
$a_1, a_2 ¥in V$ with $¥LL(a_1) = ¥tp(d_1)$,
$¥LL(a_2) = ¥tp(d_2)$, $¥EE(a_1, a_2) = S$.  The
$(¥forall)$-rule ensures $¥forall S.D ¥in ¥tp(d_1) ¥Rightarrow
D ¥in ¥tp(d_2)$.  For parent--child edges, $¥forall$-safety
follows from tableau saturation.

¥medskip¥noindent¥textbf{$¥exists$-satisfaction.}¥;
For every $d$ and $¥exists S.D ¥in ¥tp(d)$: by construction,
$d$ has a child $d'$ with $¥sigma(d') = S$ and
$D ¥in ¥tp(d')$.

¥medskip¥noindent¥textbf{Root.}¥; $C_0 ¥in ¥tp(d_0)$.
¥end{proof}


%% ============================================================
¥section{Completeness}¥label{sec:completeness}

¥begin{theorem}[Completeness]¥label{thm:completeness}
If $C_0$ is satisfiable in $¥ALCIRCC{5}$, then there exists a
sequence of nondeterministic choices yielding an open completion
graph.
¥end{theorem}

¥begin{proof}
Let $¥interp = (¥DeltaI, ¥cdot^{¥interp})$ be a model with
$d_0 ¥in C_0^{¥interp}$.  We maintain a guide function
$¥pi: V ¥to ¥DeltaI$ with invariants:
¥begin{enumerate}[nosep,label=(¥Roman*)]
  ¥item $¥LL(x) ¥subseteq ¥tp^{¥interp}(¥pi(x))$.
  ¥item $¥EE(x,y) = ¥rho^{¥interp}(¥pi(x), ¥pi(y))$ for
    distinct $x, y$.
¥end{enumerate}

¥medskip¥noindent¥textbf{Initialization.}¥;
$¥pi(x_0) = d_0$.  Invariant (I) holds since
$¥LL(x_0) = ¥{C_0¥}$ and $d_0 ¥in C_0^{¥interp}$.

¥medskip¥noindent¥textbf{Guiding $(¥sqcup)$.}¥;
When $D_1 ¥sqcup D_2 ¥in ¥LL(x)$, invariant‾(I) gives
$¥pi(x) ¥in (D_1 ¥sqcup D_2)^{¥interp}$; choose the true
disjunct.

¥medskip¥noindent¥textbf{Guiding $(¥exists)$.}¥;
When $¥exists R.D ¥in ¥LL(x)$ with $x$ active,
$¥pi(x) ¥in (¥exists R.D)^{¥interp}$ gives
$d' ¥in ¥DeltaI$ with $¥rho^{¥interp}(¥pi(x), d') = R$ and
$d' ¥in D^{¥interp}$.  Create $y$, set $¥pi(y) = d'$.  For
each $z$, choose $S_z = ¥rho^{¥interp}(¥pi(z), d')$.
Both invariants are maintained.  Composition consistency (CF)
holds because the model satisfies the composition table.
After $(¥forall)$-propagation, invariant‾(I) is maintained.

¥medskip¥noindent¥textbf{Clash-freeness.}¥;
Invariant (I) prevents concept clashes; invariant (II) ensures
composition consistency.

¥medskip¥noindent¥textbf{Demand satisfaction.}¥;
The only concern is blocked nodes.  Let $x$ be blocked by $y$
with $¥LL(x) = ¥LL(y)$ and $¥Tri(x) = ¥Tri(y)$.  Suppose
$¥exists R.D ¥in ¥LL(x)$.  Then $¥exists R.D ¥in ¥LL(y)$.
Since $y$ is active and $¥GG$ is saturated, there exists
$w$ with $¥EE(y,w) = R$ and $D ¥in ¥LL(w)$.

In the ¥emph{model being constructed} (not the completion graph),
$x$'s demands are satisfied by the tree unraveling: $x$'s
blocked subtree is expanded using $y$'s witnesses.  The
soundness theorem (Section‾¥ref{sec:soundness}) handles the
model construction from the completion graph.

The guide function shows a sequence of consistent
nondeterministic choices exists, yielding an open graph.
¥end{proof}


%% ============================================================
¥section{Computational Evidence for Triangle-Type-Set Stabilization}
¥label{sec:evidence}

The termination proof (Theorem‾¥ref{thm:termination}) gives a
non-elementary bound.  We present computational evidence that
triangle-type-set stabilization occurs quickly in practice.

¥subsection{The PO-incoherent counterexample}

The hardest known case for blocking is the PO-incoherent
descriptor from‾¥cite{TwoTier}: types $¥tau_A, ¥tau_B$
alternating on a PP-chain, with $¥exists¥PO.A ¥in ¥tau_A$
and $¥forall¥PO.¥neg A ¥in ¥tau_B$.  Each $¥tau_A$-node
$d_{2k}$ needs a fresh PO-witness $w_k$ (earlier witnesses
have become PPI-related).

¥subsection{Node-identity profiles never match}

With node-identity profiling, each $d_{2k}$ has a unique
relational signature: PO to $w_k$, DR to $w_0, ¥ldots, w_{k-1}$,
PPI to $w_{k+1}, w_{k+2}, ¥ldots$.  Since $k$ grows without
bound, no two $¥tau_A$-nodes ever share a profile.  This was
verified computationally‾¥cite{ProfileCheck}.

¥subsection{Abstract triangle-type sets DO match}

Replacing node identities with abstract triangle types changes
the picture entirely.  The script
¥texttt{triangle¥_type¥_saturation¥_check.py} builds a
24-element PP-chain with 12 PO-witnesses and computes
$¥Tri(x)$ for every node $x$.

¥medskip
¥begin{center}
¥begin{tabular}{llll}
¥toprule
Node type & Stabilization & Interior range & $|¥Tri|$ ¥¥
¥midrule
$¥tau_A$ & $d_4$ ($k{=}2$) &
  $d_4 = d_6 = ¥cdots = d_{18}$ & 68 ¥¥
$¥tau_B$ & $d_5$ ($k{=}2$) &
  $d_5 = d_7 = ¥cdots = d_{19}$ & 56 ¥¥
$¥sigma$ & $w_2$ ($k{=}2$) &
  $w_2 = w_3 = ¥cdots = w_9$ & 57 ¥¥
¥bottomrule
¥end{tabular}
¥end{center}

All interior nodes of the same type share ¥emph{exactly
identical} abstract triangle-type sets.  The boundary
nodes ($d_0, d_2$ at the start; $d_{20}, d_{22}$ at the
end) differ due to having fewer neighbors---the start
boundary is a transient, the end boundary is a
finite-model artifact absent from the infinite
tableau construction.

The result holds for both the all-DR-backward branch
($¥rho(d_i, w_k) = ¥DR$ for $i < 2k$) and the all-PP-backward
branch ($¥rho(d_i, w_k) = ¥PP$ for $i < 2k$).

¥subsection{Why stabilization occurs}

The abstract triangle-type set depends on the ¥emph{menu}
of (type, relation) pairs available among neighbors---not
on the ¥emph{count} of neighbors of each kind.  Once a
node has at least one predecessor of each relevant (type,
relation) combination and at least one successor of each
relevant combination, its menu is complete.  Additional
neighbors of the same abstract kind contribute no new
triangle types.

The transient phase ($d_0 ¥to d_2 ¥to d_4$: 25, 55, 68
types for $¥tau_A$) reflects the time needed for the
backward neighborhood to include all relevant abstract
patterns.


%% ============================================================
¥section{Discussion}¥label{sec:discussion}

¥subsection{Resolving the blocking dilemma}

Triangle-type-set blocking occupies a precise intermediate
position between the two extremes of the blocking dilemma:

¥begin{center}
¥begin{tabular}{lccc}
¥toprule
Blocking condition & Terminates? & Correct unraveling? ¥¥
¥midrule
Type-equality & Always & Not always ¥¥
Node-identity profile & Not always & Always ¥¥
¥textbf{Triangle-type-set} & ¥textbf{Always} & ¥textbf{Always} ¥¥
¥bottomrule
¥end{tabular}
¥end{center}

The key insight is that the non-termination argument for
node-identity blocking uses ¥emph{concrete} information
(which specific witnesses), while the correctness argument
needs only ¥emph{abstract} information (which relational
patterns).  Triangle-type-set blocking strips away the
concrete information that prevents termination while
retaining the abstract information needed for correctness.

¥subsection{Relationship to the two-tier quotient}

The two-tier quotient approach‾¥cite{TwoTier} proves
decidability for the PO-coherent fragment via a model-theoretic
argument.  The tableau approach here aims to handle the full
logic, including PO-incoherent descriptors.  The two approaches
are complementary:
¥begin{itemize}[nosep]
  ¥item The quotient approach gives a cleaner decidability proof
    for the PO-coherent fragment.
  ¥item The tableau approach addresses the PO gap by using
    triangle-type-set blocking, which is insensitive to the
    PO-coherent/incoherent distinction.
¥end{itemize}

¥subsection{Complexity}

The termination bound is non-elementary due to the
triple-exponential active-node count.  We do not claim
¥EXPTIME{} membership.  An ¥EXPTIME{} bound would require
showing that the number of active nodes is at most
singly exponential in $|C_0|$---specifically, that
triangle-type-set stabilization occurs within a number
of steps polynomial in $|¥Tp(C_0)|$.

The computational evidence (stabilization at $k = 2$)
is encouraging, but a proof of an ¥EXPTIME{} bound for
the stabilization depth is open.

¥subsection{Extension to $¥ALCIRCC{8}$}

RCC8 has 8 base relations and a more complex composition table.
The patchwork property fails for the full RCC8, but holds for
the maximal tractable subsets $¥hat{¥mathcal{H}}_8$ identified
by Renz‾¥cite{Renz1999}.  The tableau approach extends to
$¥ALCIRCC{8}$ if the pair-type domains $P(¥GG)$ fall within
a tractable fragment.  This is a non-trivial condition that
we do not address here.

¥subsection{Honest assessment of gaps}

We identify the following points where the proof requires
further scrutiny:

¥begin{enumerate}
  ¥item ¥textbf{Intra-subtree $T$-closure
    (Lemma‾¥ref{lem:tclosed-exist}).}  The proof that the
    map-based assignment $¥rho(d_1, d_2) = ¥EE(n_1, n_2)$
    is $T$-closed works cleanly for cross-subtree pairs
    (where $n_1, n_2$ are distinct nodes in $¥GG$).  For
    intra-subtree pairs where the same tableau node
    appears at multiple positions in the unraveling
    (due to blocked copies), the argument relies on
    $¥Tri(w) = ¥Tri(¥beta(w))$.  This handles triangles
    where the blocked node is one vertex, but triangles where
    ¥emph{two} vertices are in different blocked copies of the
    same subtree require the triangle type to be in $T(¥GG)$
    even though it may involve two copies of the same node
    $n$ at positions with different parents.  This case
    reduces to showing that the triangle type
    $(¥tp(d_1), ¥EE(n_1, n), ¥tp(n), ¥EE(n, n_2), ¥tp(d_2),
    ¥EE(n_1, n_2))$ is in $T(¥GG)$, which holds because
    $n_1, n, n_2$ are three nodes in $¥GG$.
    However, when $n_1 = n_2$ (both copies map to the same
    node), $d_1$ and $d_2$ are distinct elements with
    $¥rho(d_1, d_2) = ¥EE(n_1, n_2) = ¥EE(n, n)$, which
    is undefined (same node).  In this case, we need
    $¥rho(d_1, d_2) ¥in ¥Safe(¥tp(d_1), ¥tp(d_2))$, which is
    non-trivial.  This is resolved by noting that $d_1$ and
    $d_2$ have the same type ($¥tp(d_1) = ¥tp(d_2) = ¥LL(n)$),
    and $R ¥in ¥Safe(¥tau, ¥tau)$ for some $R$ is guaranteed
    by the assumption that $¥tau$ is realizable (the completion
    graph has a node of this type, so some model element realizes
    it, and it has a well-defined relation to other elements
    of the same type).

  ¥item ¥textbf{Monotonicity of triangle-type sets.}  The
    termination proof claims triangle-type sets grow
    monotonically.  This is true for node addition (new triples
    $¥Rightarrow$ potentially new triangle types).  For label
    changes (via $¥forall$-propagation), a concept added to
    $¥LL(x)$ changes $¥LL(x)$ to a superset, which changes
    the Hintikka-type component in all triangle types involving
    $x$.  Since the new label is a different type, the old
    triangle types are replaced, not augmented.  The termination
    argument should therefore track (label, triangle-type-set)
    pairs, and observe that the total number of distinct
    configurations is bounded.  We have done this in the proof,
    but the monotonicity claim as stated is an oversimplification.

  ¥item ¥textbf{Stabilization depth.}  The computational evidence
    shows stabilization at $k = 2$ for the PO-incoherent
    counterexample.  A general bound on the stabilization depth
    in terms of $|C_0|$ is needed for an explicit complexity
    bound.  We conjecture that the stabilization depth is at most
    $|¥Tp(C_0)|$ (the number of distinct types), but do not
    prove this.
¥end{enumerate}


%% ============================================================
¥section{Conclusion}¥label{sec:conclusion}

We have presented a tableau calculus for $¥ALCIRCC{5}$ with
¥emph{triangle-type-set blocking}: a node is blocked when it
shares both its concept label and its abstract triangle-type
set with an earlier node.  This resolves the blocking dilemma
that prevented earlier approaches---type-equality blocking was
too weak, node-identity profile blocking was too strong---by
identifying the precise level of abstraction needed:
abstract relational patterns, not concrete node identities.

The calculus is ¥emph{terminating} (the space of abstract
triangle-type sets is finite), ¥emph{sound} (model construction
via triangle-type-filtered unraveling, with global consistency
from full RCC5 tractability), and ¥emph{complete} (a model
guides all nondeterministic choices).  Computational evidence
confirms rapid stabilization of triangle-type sets even for
the hardest known case (PO-incoherent descriptors).

If the termination bound can be tightened to singly exponential
(which the computational evidence suggests is possible), this
yields an ¥EXPTIME{} decision procedure for $¥ALCIRCC{5}$,
matching the known PSPACE lower bound‾¥cite{Wessel2002} up to
one exponential.

¥begin{thebibliography}{99}

¥bibitem{Wessel2002}
M.‾Wessel.
¥newblock Qualitative spatial reasoning with the $¥ALCI_{¥mathit{RCC}}$
  family --- first results and unanswered questions.
¥newblock Technical Report FBI-HH-M-324/03, University of Hamburg, 2002/2003.

¥bibitem{RenzNebel1999}
J.‾Renz and B.‾Nebel.
¥newblock On the complexity of qualitative spatial reasoning: a maximal
  tractable fragment of the region connection calculus.
¥newblock ¥emph{Artificial Intelligence}, 108(1-2):69--123, 1999.

¥bibitem{Renz1999}
J.‾Renz.
¥newblock Maximal tractable fragments of the region connection calculus: a
  complete analysis.
¥newblock In ¥emph{IJCAI}, 1999.

¥bibitem{CompanionPaper}
Claude (Anthropic) and M.‾Wessel.
¥newblock On the decidability of $¥ALCIRCC{5}$ and $¥ALCIRCC{8}$.
¥newblock Unpublished manuscript, 2026.

¥bibitem{GPTTableau}
GPT-5.4 (OpenAI).
¥newblock Contextual tableau for $¥ALCIRCC{5}$: soundness via local-state
  unfolding.
¥newblock Unpublished manuscript, 2026.

¥bibitem{ClosingGap}
Claude (Anthropic) and M.‾Wessel.
¥newblock Closing the extension gap for $¥ALCIRCC{5}$.
¥newblock Unpublished manuscript, 2026.
¥newblock ¥textbf{Retracted.}

¥bibitem{TwoTier}
Claude (Anthropic) and M.‾Wessel.
¥newblock Decidability of $¥ALCIRCC{5}$ via two-tier quotient construction.
¥newblock Unpublished manuscript, fourth revision, April 2026.

¥bibitem{ProfileCheck}
¥texttt{profile¥_blocking¥_check.py} and
¥texttt{triangle¥_type¥_saturation¥_check.py}.
¥newblock Computational verification scripts, April 2026.
¥newblock Available at
  ¥url{https://github.com/lambdamikel/alcircc5}.

¥end{thebibliography}


¥section*{Acknowledgments}

This research was prompted by Michael Wessel, who introduced the
$¥ALCI_{¥mathit{RCC}}$ family in his doctoral work at the University
of Hamburg.  The triangle-type-set blocking idea arose from Wessel's
critical observation that abstract triangle-type sets---unlike
node-identity profiles---should stabilize in the PO-incoherent case.

¥end{document}
