\documentclass[a4paper,11pt]{article}

%%% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[margin=2.5cm]{geometry}
\usepackage{xcolor}

\hypersetup{colorlinks=true,linkcolor=blue!45!black,citecolor=blue!45!black,urlcolor=blue!45!black}

%%% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{claim}[theorem]{Claim}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

%%% Notation macros
\newcommand{\ALCI}{\ensuremath{\mathcal{ALCI}}}
\newcommand{\ALCIRCC}[1]{\ensuremath{\mathcal{ALCI}_{\mathit{RCC#1}}}}
\newcommand{\DR}{\ensuremath{\mathsf{DR}}}
\newcommand{\PO}{\ensuremath{\mathsf{PO}}}
\newcommand{\EQ}{\ensuremath{\mathsf{EQ}}}
\newcommand{\PP}{\ensuremath{\mathsf{PP}}}
\newcommand{\PPI}{\ensuremath{\mathsf{PPI}}}
\newcommand{\inv}{\ensuremath{\mathsf{inv}}}
\newcommand{\comp}{\ensuremath{\mathsf{comp}}}
\newcommand{\cl}{\ensuremath{\mathsf{cl}}}
\newcommand{\tp}{\ensuremath{\mathsf{tp}}}
\newcommand{\Tp}{\ensuremath{\mathsf{Tp}}}
\newcommand{\EXPTIME}{\textsc{ExpTime}}
\newcommand{\Desc}{\ensuremath{\mathsf{Desc}}}
\newcommand{\Core}{\ensuremath{\mathsf{Core}}}
\newcommand{\interp}{\ensuremath{\mathcal{I}}}
\newcommand{\DeltaI}{\ensuremath{\Delta^{\mathcal{I}}}}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0.55em}


\title{\textbf{Decidability of $\ALCIRCC{5}$ via\\Two-Tier Quotient Construction}\\[0.4em]
\large Period descriptors, PP-kernel nodes,\\
and the full tractability of RCC5}
\author{Claude (Anthropic)\\
\small Prompted by Michael Wessel}
\date{April 2026}

\begin{document}
\maketitle

\begin{center}
\fbox{\parbox{0.92\textwidth}{%
\textbf{Disclaimer.}\; This document was generated by Claude
(Anthropic), prompted by Michael Wessel.  The results have not been
independently verified.  We invite scrutiny and corrections.}}
\end{center}

\begin{abstract}
We prove that concept satisfiability in $\ALCIRCC{5}$---the
description logic $\ALCI$ extended with an RCC5 role
box---is decidable.  The proof proceeds by a model-theoretic quotient
construction.  Every model of a satisfiable concept can be
represented by a finite \emph{two-tier quotient}: \textbf{Tier~1}
captures each infinite $\PP$-chain's eventually periodic tail as a
finite \emph{period descriptor} (a cyclic word of Hintikka types),
while \textbf{Tier~2} captures cross-chain interactions through
\emph{PP-kernel nodes} with reflexive $\PP$-loops.  Conversely, every
valid two-tier quotient can be unfolded into a genuine model, using
the \emph{full tractability} of RCC5 (Renz \& Nebel 1999): path-consistent
disjunctive RCC5 networks are satisfiable.  Together, these yield a
decision procedure: enumerate quotients up to a computable size bound
and check validity.  The key algebraic ingredient is that reflexive
$\PP$ is \emph{universally self-absorbing}, ensuring composition
consistency of the kernel representation.
\end{abstract}


%% ============================================================
\section{Introduction}\label{sec:intro}

The description logic $\ALCIRCC{5}$ extends $\ALCI$ with five jointly
exhaustive and pairwise disjoint (JEPD) base relations from the
Region Connection Calculus RCC5~\cite{RandellCuiCohn1992}: $\DR$,
$\PO$, $\EQ$, $\PP$, $\PPI$.  Wessel~\cite{Wessel2003} introduced
this logic and left its decidability open.  The logic lacks both the
finite model property and the tree model property: the concept
$C_\infty = (\exists\PP.\top) \sqcap (\forall\PP.\exists\PP.\top)$ is
satisfiable only in infinite models containing unbounded $\PP$-chains.

Several recent approaches have been attempted:
a quasimodel method~\cite{Claude2026} (gap in completeness),
a contextual tableau~\cite{GPT2026} (whose completeness conjecture
$\mathsf{FW}(C,N)$ is false~\cite{ClaudeFW2026}), and a
triangle-type approach~\cite{ClaudeTriangle2026} (conditional on an
unproven conjecture).  All share a common difficulty: assigning
globally consistent RCC5 edges in complete-graph models.

In this paper we resolve the decidability question affirmatively.
The proof has two parts:

\textbf{Completeness} (satisfiable concept $\Rightarrow$ finite
quotient): Given any model $\interp \models C_0$, we extract a
finite \emph{two-tier quotient}.  Each infinite $\PP$-chain is
represented by a \emph{period descriptor}---a cyclic word
$(\tau_1, \ldots, \tau_p)$ of Hintikka types capturing the chain's
eventually periodic structure.  Cross-chain interactions are
represented by \emph{PP-kernel nodes} with reflexive $\PP$-loops
and stabilized inter-chain relations.  The model's global
consistency guarantees that the quotient satisfies a checkable
\emph{extension consistency} condition (V6).

\textbf{Soundness} (valid quotient $\Rightarrow$ satisfiable
concept): Given a valid two-tier quotient, we unfold it into a
model.  Period descriptors become infinite $\PP$-chains.
Off-chain witnesses are added via an extension procedure whose
consistency is guaranteed by V6 and the \emph{full tractability of
RCC5}~\cite{RenzNebel1999}: path-consistent disjunctive RCC5
constraint networks are satisfiable.

The decision procedure enumerates all two-tier quotients up to a
computable size bound and checks validity.  The quotient size is
bounded by a computable (though possibly non-elementary) function of
the input concept.


%% ============================================================
\section{Preliminaries}\label{sec:prelim}

\subsection{The logic $\ALCIRCC{5}$}

Concepts are built from concept names, the RCC5 base relations
$R \in \{\DR, \PO, \EQ, \PP, \PPI\}$ as role names, and the
constructors $\sqcap$, $\sqcup$, $\neg$, $\exists R.C$,
$\forall R.C$ in negation normal form (NNF).  We adopt
\textbf{strong $\EQ$ semantics}.

An interpretation $\interp = (\DeltaI, \cdot^{\interp})$ assigns a
total function $\rho^{\interp}: \DeltaI \times \DeltaI \to
\{\DR, \PO, \EQ, \PP, \PPI\}$ subject to: (i)~$\rho(d,d) = \EQ$;
(ii)~$\rho(d,e) = \inv(\rho(e,d))$; (iii)~composition consistency:
$\rho(d_1, d_3) \in \comp(\rho(d_1, d_2), \rho(d_2, d_3))$ for all
$d_1, d_2, d_3$.

\subsection{Types, closure, composition table}

Given $C_0$ in NNF, $\cl(C_0)$ is the closure under subformulas.
A \emph{Hintikka type} $\tau \subseteq \cl(C_0)$ is a maximal
propositionally consistent subset.  Write $\Tp(C_0)$ for the set of
types; $|\Tp(C_0)| \le 2^{|\cl(C_0)|}$.

The RCC5 composition table
$\comp: \{\DR,\PO,\PP,\PPI\}^2 \to 2^{\{\DR,\PO,\PP,\PPI\}}$ (with
$\EQ$ excluded for distinct elements) is:

\begin{center}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{c|cccc}
$\comp(R_1, R_2)$ & $\DR$ & $\PO$ & $\PP$ & $\PPI$ \\
\hline
$\DR$ & \{all\} & \{\DR,\PO,\PP\} & \{\DR,\PO,\PP\} & \{\DR\} \\
$\PO$ & \{\DR,\PO,\PPI\} & \{all\} & \{\PO,\PP\} & \{\DR,\PO,\PPI\} \\
$\PP$ & \{\DR\} & \{\DR,\PO,\PP\} & \{\PP\} & \{all\} \\
$\PPI$ & \{\DR,\PO,\PPI\} & \{\PO,\PPI\} & \{\PO,\PP,\PPI\} & \{\PPI\} \\
\end{tabular}
\end{center}

\subsection{Full tractability of RCC5}

The following deep result from qualitative spatial reasoning is
central to our proof.

\begin{theorem}[Full RCC5 tractability, Renz \& Nebel
\protect{\cite{RenzNebel1999}}, Renz
\protect{\cite{Renz1999}}]\label{thm:full-tractability}
Every path-consistent disjunctive RCC5 constraint network over the
base relations $\{\DR, \PO, \EQ, \PP, \PPI\}$ is satisfiable.
Equivalently: if path-consistency enforcement (= arc-consistency for
the composition constraints) does not empty any domain, then a
consistent atomic refinement exists.\qed
\end{theorem}

This is the ``patchwork property'' of RCC5: local consistency
(path-consistency) implies global consistency.  For atomic networks:
every path-consistent atomic RCC5 network is realizable by regions in
$\mathbb{R}^n$ for sufficiently large $n$.


%% ============================================================
\section{Algebraic properties of $\PP$}\label{sec:pp-algebra}

\begin{theorem}[Universal self-absorption of $\PP$]
\label{thm:self-absorbing}
\begin{enumerate}[label=(\alph*)]
\item \textbf{Left absorption:} $R \in \comp(\PP, R)$ for every
  $R \in \{\DR,\PO,\PP,\PPI\}$.
\item \textbf{Right absorption:} $R \in \comp(R, \PP)$ for every $R$.
\item \textbf{Self-closure:} $\PP \in \comp(R, \inv(R))$ for every $R$.
\item \textbf{PPI right-absorption:} $R \in \comp(\PPI, R)$ for every $R$.
\item \textbf{PPI left-absorption:} $R \in \comp(R, \PPI)$ for every $R$.
\end{enumerate}
\end{theorem}

\begin{proof}
Direct verification against the composition table.  Parts (d)--(e)
are the dual of (a)--(b) under the involution $\PP \leftrightarrow
\PPI$.
\end{proof}

\begin{corollary}[Reflexive $\PP$ is composition-consistent]
\label{cor:reflexive-pp}
A node $k$ with $\rho(k,k) = \PP$ satisfies all composition
constraints with any external node $m$.\qed
\end{corollary}

\begin{theorem}[$\PP$-transitivity and strict order]
\label{thm:no-pp-cycle}
$\comp(\PP, \PP) = \{\PP\}$.  Consequently, $\PP$ on any finite set
of elements is a strict partial order, and no $\PP$-cycle exists among
distinct elements.\qed
\end{theorem}

\begin{lemma}[Non-empty compositions]\label{lem:nonempty}
$\comp(R_1, R_2) \ne \emptyset$ for all $R_1, R_2 \in
\{\DR,\PO,\PP,\PPI\}$. \qed
\end{lemma}


%% ============================================================
\section{$\PP$-chain structure}\label{sec:chain}

\begin{definition}[$\PP$-chain]
In a model $\interp$, a \emph{$\PP$-chain} is a sequence
$(d_i)_{i \ge 0}$ with $\rho(d_i, d_{i+1}) = \PP$ for all $i$.
By Theorem~\ref{thm:no-pp-cycle}, $\rho(d_i, d_j) = \PP$ for $i < j$.
\end{definition}

\begin{lemma}[Monotone accumulation]\label{lem:monotone}
Define $F_i = \{C \in \cl(C_0) : \forall\PP.C \in \tp(d_i)\}$.
Then $F_0 \subseteq F_1 \subseteq \cdots$, and the sequence
stabilizes within $|\cl(C_0)|$ steps.\qed
\end{lemma}

\begin{definition}[Stabilized tail]
The \emph{stabilized tail} of a $\PP$-chain $(d_i)_{i \ge 0}$ is
$(d_i)_{i \ge N}$ where $N$ is the stabilization index.
\end{definition}

\begin{lemma}[External relation stabilization]\label{lem:external}
For external element $e$ and chain $(d_i)$, the sequence
$R_i = \rho(e, d_i)$ satisfies $R_{i+1} \in \comp(R_i, \PP)$
and stabilizes.  $\PP$ is the unique absorbing value:
$\comp(\PP, \PP) = \{\PP\}$.\qed
\end{lemma}

\begin{definition}[Recurrent types, $T_\infty$]\label{def:tinf}
For a $\PP$-chain $(d_i)_{i \ge 0}$, define
$T_\infty = \{\tau \in \Tp(C_0) : \tau = \tp(d_i) \text{ for
infinitely many } i\}$.
\end{definition}

\begin{lemma}[$T_\infty$ is non-empty and eventually
exhaustive]\label{lem:tinf}
$T_\infty \ne \emptyset$.  There exists $M \ge N$ (the stabilization
index) such that $\tp(d_i) \in T_\infty$ for all $i \ge M$.
\end{lemma}

\begin{proof}
Since $|\Tp(C_0)|$ is finite and the sequence $(\tp(d_i))_{i \ge 0}$
is infinite, some type appears infinitely often (pigeonhole), so
$T_\infty \ne \emptyset$.  Types not in $T_\infty$ appear only
finitely often; let $M$ be the last index at which any
non-$T_\infty$ type appears.  For $i > M$, $\tp(d_i) \in T_\infty$.
\end{proof}


%% ============================================================
\section{Period descriptors}\label{sec:period}

\begin{definition}[Period descriptor]\label{def:period-descriptor}
A \emph{period descriptor} for $C_0$ is a tuple
$\Desc = (\tau_1, \ldots, \tau_p)$ with $p \ge 1$ and each
$\tau_j \in \Tp(C_0)$, satisfying:
\begin{enumerate}[label=(V\arabic*)]
\item Each $\tau_j$ is a Hintikka type in $\cl(C_0)$.
\item \textbf{$\forall\PP$-closure:} $\forall\PP.C \in \tau_i$
  for some $i$ implies $C \in \tau_j$ for all $j$.
\item \textbf{$\forall\PPI$-closure:} $\forall\PPI.C \in \tau_i$
  for some $i$ implies $C \in \tau_j$ for all $j$.
\item \textbf{Chain continuation:} For each $i$, there exists
  $\exists\PP.C \in \tau_i$ such that $C \in \tau_j$ for some~$j$.
\end{enumerate}
The descriptor represents the infinite periodic chain
$\cdots \tau_1 \;\PP\; \tau_2 \;\PP\; \cdots \;\PP\; \tau_p
\;\PP\; \tau_1 \;\PP\; \cdots$
\end{definition}

\begin{remark}[V4 is chain continuation, not full witness closure]
Condition V4 requires that each type has \emph{at least one}
$\exists\PP$ demand satisfiable within the period (ensuring the chain
can continue).  Other $\exists\PP$ demands of $\tau_i$ may be
satisfied by off-chain witnesses (handled by Tier~2 of the quotient).
This weakening is essential: an element on a $\PP$-chain may have
multiple $\exists\PP$ demands, only one of which generates the next
chain element.  The others generate off-chain $\PP$-branches.
\end{remark}

\begin{theorem}[$T_\infty$ yields a valid period descriptor]
\label{thm:tinf-descriptor}
Let $(d_i)_{i \ge 0}$ be a $\PP$-chain in a model $\interp \models C_0$.
Any enumeration of $T_\infty = \{\tau_1, \ldots, \tau_p\}$ is a valid
period descriptor.
\end{theorem}

\begin{proof}
(V1): Each $\tau_j \in T_\infty \subseteq \Tp(C_0)$.

(V2): If $\forall\PP.C \in \tau_i$ for $\tau_i \in T_\infty$, then
$\tau_i$ appears at some position $k \ge M$ on the chain.  Since
$\forall\PP.C \in \tp(d_k)$ and $\rho(d_k, d_j) = \PP$ for all
$j > k$, we have $C \in \tp(d_j)$ for all $j > k$.  In particular,
$C \in \tau$ for every $\tau \in T_\infty$ (since every $\tau \in
T_\infty$ appears at some position $j > k$).

(V3): Dual argument using $\forall\PPI$ and backward propagation.
If $\forall\PPI.C \in \tau_i$ at position $k$, then $C \in \tp(d_j)$
for all $j < k$ with $\rho(d_k, d_j) = \PPI$.  On the chain,
$\rho(d_k, d_j) = \PPI$ for $j < k$.  So $C$ is in all types
at positions $< k$.  Since $\tau_i$ appears at arbitrarily large $k$
(infinitely often), $C$ is in all types at all positions $> M$,
hence in all $\tau \in T_\infty$.

(V4): Each $d_k$ (with $k \ge M$, $\tp(d_k) \in T_\infty$) has a
$\PP$-successor $d_{k+1}$ on the chain.  The element $d_{k+1}$
witnesses some $\exists\PP.C$ demand of $d_k$, and
$\tp(d_{k+1}) \in T_\infty$.  So $C \in \tp(d_{k+1})$ for some
$\tp(d_{k+1}) \in T_\infty$.
\end{proof}

\begin{definition}[Stabilized core]
$\Core(\Desc) = \bigcap_{j=1}^{p} \tau_j$ for
$\Desc = (\tau_1, \ldots, \tau_p)$.
By (V2) and (V3), $\Core(\Desc)$ contains all $\forall\PP$ and
$\forall\PPI$ consequences.
\end{definition}


%% ============================================================
\section{The two-tier quotient}\label{sec:quotient}

\subsection{Witness trees and PP-chains}

Fix a model $\interp \models C_0$.  The \emph{tree unraveling}
$\interp^T$ of $\interp$ is a tree-shaped model where each node is a
domain element and each edge corresponds to an existential witness.
In $\interp^T$:
\begin{itemize}
\item The root satisfies $C_0$.
\item Each node $d$ has one child for each $\exists R.C \in \tp(d)$
  (the child witnesses the demand via relation $R$).
\item A \emph{maximal $\PP$-chain} is a maximal path through the
  tree following $\exists\PP$-witness edges.
\end{itemize}

\begin{remark}
A node may have multiple $\exists\PP$ demands.  One generates the
next element on the main $\PP$-chain; the others generate
\emph{off-chain $\PP$-branches}, which are themselves roots of
sub-$\PP$-chains.  The tree of $\PP$-chains forms a recursive
structure where each chain may spawn sub-chains at each position.
\end{remark}

\subsection{Definition of the two-tier quotient}

\begin{definition}[Two-tier quotient]\label{def:quotient}
A \emph{two-tier quotient} $\mathcal{Q}$ for $C_0$ consists of:

\medskip\noindent\textbf{Tier~1 (within-chain structure):}
A finite set of period descriptors
$\{\Desc_1, \ldots, \Desc_K\}$, each satisfying (V1)--(V4).

\medskip\noindent\textbf{Tier~2 (cross-chain structure):}
\begin{enumerate}[label=(T\arabic*)]
\item A set of \emph{PP-kernel nodes}
  $\{k_1, \ldots, k_K\}$, one per descriptor, with
  $\tp(k_\alpha) = \Core(\Desc_\alpha)$ and
  $\rho(k_\alpha, k_\alpha) = \PP$.
\item A set of \emph{regular nodes} $\{m_1, \ldots, m_L\}$ with
  types $\tp(m_\beta) \in \Tp(C_0)$.
\item An atomic RCC5 labeling $\rho$ on all pairs of
  kernel and regular nodes (excluding self-pairs of regular nodes,
  which have $\rho = \EQ$), satisfying composition consistency for
  every triple.
\item For each kernel node $k_\alpha$ and each non-$\PP$ existential
  demand $\exists R.C \in \Core(\Desc_\alpha)$ with
  $R \ne \PP$: a designated witness among the kernel/regular nodes.
\item For each regular node $m_\beta$ and each existential demand
  $\exists R.C \in \tp(m_\beta)$: a designated witness.
\end{enumerate}

\medskip\noindent\textbf{Extension consistency:}
\begin{enumerate}[label=(V\arabic*),start=5]
\setcounter{enumi}{4}
\item \textbf{Cross-chain coherence:} The atomic network on
  kernel and regular nodes is composition-consistent.
\item \textbf{Off-chain extension:} For each $\exists\PP.C$ demand
  of a kernel node $k_\alpha$ that is \emph{not} satisfied within
  $\Desc_\alpha$ (i.e., $C \notin \tau_j$ for any $j$), there is
  a witness node $w$ with $\rho(k_\alpha, w) = \PP$ and
  $C \in \tp(w)$.  Furthermore, the disjunctive constraint network
  obtained by adding all such off-chain witnesses simultaneously
  (with demanded relations to their parents and initial domains
  $D(w, e) = \comp(\text{demanded}, \rho(\text{parent}, e))$
  for all existing nodes $e$) is \emph{path-consistent} after
  arc-consistency enforcement (i.e., no domain becomes empty).
\end{enumerate}
\end{definition}

\begin{remark}
The off-chain witnesses in V6 may themselves have existential demands
that require further witnesses, forming a tree of extensions.  V6
checks the \emph{joint} path-consistency of all such extensions up to
a blocking depth (see Section~\ref{sec:completeness}).  This is a
finite computation.
\end{remark}


%% ============================================================
\section{Completeness: model implies quotient}
\label{sec:completeness}

\begin{theorem}[Quotient extraction]\label{thm:completeness}
If $C_0$ is satisfiable, there exists a valid two-tier quotient
$\mathcal{Q}$ for $C_0$ of size bounded by a computable function
$f(|C_0|)$.
\end{theorem}

\begin{proof}
Let $\interp \models C_0$.  Form the tree unraveling $\interp^T$.

\medskip\noindent\textit{Step 1: Period descriptors.}\;
For each maximal $\PP$-chain in $\interp^T$, compute $T_\infty$
(Definition~\ref{def:tinf}).  By Theorem~\ref{thm:tinf-descriptor},
any enumeration of $T_\infty$ is a valid period descriptor satisfying
(V1)--(V4).  The number of distinct descriptors is at most
$2^{|\Tp(C_0)|} \le 2^{2^{|\cl(C_0)|}}$ (bounded by the number of
subsets of $\Tp(C_0)$).

\medskip\noindent\textit{Step 2: Kernel nodes and stabilized
relations.}\;
For each distinct descriptor $\Desc_\alpha$, create kernel node
$k_\alpha$ with $\tp(k_\alpha) = \Core(\Desc_\alpha)$ and
$\rho(k_\alpha, k_\alpha) = \PP$.

For two kernel nodes $k_\alpha, k_\beta$ representing chains
$A = (a_i)$ and $B = (b_j)$: define $\rho(k_\alpha, k_\beta)$ as
the \emph{doubly stabilized} relation: $\rho(a_i, b_j)$ for
sufficiently large $i, j$.  This stabilizes by
Lemma~\ref{lem:external} applied twice (first fix $j$, stabilize in
$i$; then stabilize the result in $j$).

\medskip\noindent\textit{Step 3: Composition consistency of the
kernel network.}\;
The kernel-level relations are stabilized model relations.  For any
triple $(k_\alpha, k_\beta, k_\gamma)$: choose $i, j, \ell$
large enough that all three pairwise relations have stabilized.  The
triple $(a_i, b_j, c_\ell)$ in the model is composition-consistent
(the model is globally consistent).  The stabilized relations equal
the model relations at these indices, hence (V5) holds.

Similarly for triples involving kernel nodes and regular nodes
(regular nodes have their original model relations).

\medskip\noindent\textit{Step 4: Regular nodes.}\;
The finite prefix of each $\PP$-chain (before the stabilized tail)
and all off-chain witnesses (up to a blocking depth) are recorded as
regular nodes.  Two off-chain witnesses are \emph{blocked} (identified)
if they share:
\begin{itemize}
\item the same Hintikka type $\tau$,
\item the same demanded relation $R$ to their parent,
\item the same stabilized relations to all kernel nodes.
\end{itemize}
The number of blocking classes is at most
$|\Tp(C_0)| \times 4 \times 4^K$ where $K$ is the number of kernels.
The tree of off-chain witnesses has depth bounded by the number of
blocking classes (when a class repeats on a branch, the subtree below
the repeat can be redirected to the earlier occurrence).  So the
number of regular nodes is bounded by a computable function of
$|C_0|$.

\medskip\noindent\textit{Step 5: Extension consistency (V6).}\;
Each off-chain witness $w$ in the model has actual RCC5 relations to
all other elements.  We show the model's assignment survives
arc-consistency enforcement.

For any existing element $e$ in the quotient, the model gives
$\rho^{\interp}(w, e)$.  This value satisfies:
\begin{enumerate}[label=(\roman*)]
\item $\rho(w, e) \in \comp(\rho(w, \text{parent}),
  \rho(\text{parent}, e))$ (composition in the model);
\item $\rho(w, e) \in \comp(\rho(w, e'), \rho(e', e))$ for every
  other element $e'$ (global consistency of the model).
\end{enumerate}

Arc-consistency enforcement removes a value $r$ from domain $D(w, e)$
only if there exists $e'$ such that $\comp(r, \rho(e, e')) \cap
D(w, e') = \emptyset$.  But $\rho(w, e')$ is in $D(w, e')$
(by the same argument) and $\rho(w, e') \in \comp(\rho(w, e),
\rho(e, e'))$ (global consistency), so $\comp(\rho(w, e), \rho(e,
e')) \cap D(w, e') \ni \rho(w, e') \ne \emptyset$.  Therefore
$\rho(w, e)$ is never removed.  All domains remain non-empty.
\end{proof}

\begin{remark}[The critical step]
The proof of V6 (Step~5) is where the global consistency of the
model is used most essentially.  The model provides a specific
atomic assignment for each off-chain witness that is simultaneously
consistent with \emph{all} composition constraints.  Arc-consistency
enforcement cannot eliminate a value that participates in a globally
consistent assignment.  This is the key insight that closes the
``extension gap'' identified in earlier approaches.
\end{remark}


%% ============================================================
\section{Soundness: quotient implies model}\label{sec:soundness}

\begin{theorem}[Quotient realization]\label{thm:soundness}
If a valid two-tier quotient $\mathcal{Q}$ exists for $C_0$ (i.e.,
$\mathcal{Q}$ satisfies V1--V6), then $C_0$ is satisfiable.
\end{theorem}

\begin{proof}
We construct a model $\interp' \models C_0$ from $\mathcal{Q}$.

\medskip\noindent\textit{Step 1: Unfold period descriptors.}\;
For each descriptor $\Desc_\alpha = (\tau_1, \ldots, \tau_p)$,
create an infinite $\PP$-chain
$d_1^{(\alpha)}, d_2^{(\alpha)}, d_3^{(\alpha)}, \ldots$ with
$\tp(d_i^{(\alpha)}) = \tau_{((i-1) \bmod p) + 1}$ and
$\rho(d_i^{(\alpha)}, d_j^{(\alpha)}) = \PP$ for $i < j$.

\medskip\noindent\textit{Step 2: Assign inter-chain and
chain-to-regular relations.}\;
For chain elements $d_i^{(\alpha)}$ (on the periodic tail, $i$
large enough) and any other node $e$: set
$\rho(d_i^{(\alpha)}, e) = \rho(k_\alpha, e)$, the kernel's
relation.

This is well-defined because the kernel's relation is the stabilized
value.  For chain elements in the finite prefix, use the regular
nodes' recorded relations.

\medskip\noindent\textit{Step 3: Verify within-chain
consistency.}\;
For a triple $(d_i, d_j, d_k)$ on the same chain with $i < j < k$:
$\rho(d_i, d_j) = \rho(d_j, d_k) = \PP$ and $\rho(d_i, d_k)
\in \comp(\PP, \PP) = \{\PP\}$. \checkmark

For a triple $(d_i, d_j, e)$ with $d_i, d_j$ on the same chain and
$e$ external: $\rho(d_i, d_j) = \PP$, $\rho(d_i, e) =
\rho(d_j, e) = R$ (both stabilized to $R$).  Need:
$R \in \comp(\PPI, R)$.  This holds by
Theorem~\ref{thm:self-absorbing}(d). \checkmark

For a triple $(d_i, e_1, e_2)$ with $d_i$ on a chain and $e_1, e_2$
external: $\rho(d_i, e_1) = \rho(k_\alpha, e_1)$,
$\rho(d_i, e_2) = \rho(k_\alpha, e_2)$,
$\rho(e_1, e_2)$ from the quotient.  This is exactly the triple
$(k_\alpha, e_1, e_2)$, which is composition-consistent by~(V5).
\checkmark

\medskip\noindent\textit{Step 4: Add off-chain witnesses.}\;
V6 ensures: the disjunctive network obtained by adding all off-chain
witness variables (with their demanded relations and initial domains)
to the kernel/regular network is path-consistent after arc-consistency
enforcement.

By Theorem~\ref{thm:full-tractability} (full RCC5 tractability),
this path-consistent disjunctive network is \textbf{satisfiable}: a
consistent atomic refinement exists.  Choose such a refinement.
Each off-chain witness $w$ receives specific atomic relations
$\rho(w, e)$ to all other nodes.

\medskip\noindent\textit{Step 5: Demand satisfaction.}\;
\begin{itemize}
\item \textbf{$\exists\PP$ demands within chains:} For each
  $\tau_i$ in descriptor $\Desc_\alpha$ and each $\exists\PP.C \in
  \tau_i$: if $C \in \tau_j$ for some $j$ (within-period witness),
  then the element at the next occurrence of $\tau_j$ on the chain
  witnesses the demand ($\rho = \PP$ by transitivity).  If $C$ is
  not in any period type, the demand is witnessed by an off-chain
  $\PP$-witness (from V6).

\item \textbf{Non-$\PP$ demands of chain elements:} For
  $\exists R.C \in \Core(\Desc_\alpha)$ with $R \ne \PP$: the
  quotient provides a designated witness (T2, item~4).

\item \textbf{Demands of regular nodes:} The quotient provides
  designated witnesses (T2, item~5).

\item \textbf{Universal constraints:} For $\forall R.C \in \tau_i$:
  all $R$-neighbors must have $C$ in their type.  Within chains, this
  follows from (V2) for $R = \PP$ and (V3) for $R = \PPI$.  For
  $R \in \{\DR, \PO\}$, no chain element is $\DR$ or $\PO$ from
  another chain element on the same chain (they are all $\PP$ or
  $\PPI$).  Cross-chain universal constraints are satisfied by the
  type assignments in the quotient.
\end{itemize}

\medskip\noindent\textit{Step 6: Global consistency.}\;
The full atomic network (chains + off-chain witnesses + regular nodes)
is composition-consistent:
\begin{itemize}
\item Within-chain triples: Step~3.
\item Triples involving one chain element and two
  kernel/regular/off-chain nodes: equivalent to the corresponding
  kernel triple (by stabilization) or covered by the V6 solution.
\item Triples of off-chain witnesses: consistent by the atomic
  refinement from Step~4.
\end{itemize}

By the patchwork property, this composition-consistent atomic network
is globally consistent and realizable.  The realization, combined with
the type assignments, gives a valid $\ALCIRCC{5}$ model
$\interp' \models C_0$.
\end{proof}


%% ============================================================
\section{Main result: decidability}\label{sec:decidability}

\begin{theorem}[Decidability of $\ALCIRCC{5}$]
\label{thm:decidability}
Concept satisfiability in $\ALCIRCC{5}$ is decidable.
\end{theorem}

\begin{proof}
Given $C_0$:
\begin{enumerate}
\item Compute the bound $f(|C_0|)$ on quotient size
  (Theorem~\ref{thm:completeness}).
\item Enumerate all two-tier quotient structures of size
  $\le f(|C_0|)$.  This is a finite set: finitely many period
  descriptors (sequences of types of bounded length), finitely many
  kernel/regular configurations, finitely many atomic RCC5 labelings.
\item For each quotient structure, check validity: (V1)--(V4) for
  each descriptor, (V5) composition consistency of the kernel/regular
  network, (V6) path-consistency of the off-chain extension.  Each
  check is decidable (finite computation).
\item Accept iff any valid quotient is found.
\end{enumerate}

Correctness:
\begin{itemize}
\item \emph{No false negatives:} If $C_0$ is satisfiable, then by
  Theorem~\ref{thm:completeness}, a valid quotient of size
  $\le f(|C_0|)$ exists, so the procedure accepts.
\item \emph{No false positives:} If a valid quotient exists, then
  by Theorem~\ref{thm:soundness}, $C_0$ is satisfiable.
\end{itemize}

The enumeration is finite and each check is decidable, so the
procedure terminates.
\end{proof}


%% ============================================================
\section{Complexity}\label{sec:complexity}

The quotient size bound $f(|C_0|)$ is large:
\begin{itemize}
\item Period descriptors: length $\le |\Tp(C_0)| \le 2^{|\cl(C_0)|}$.
  Number of distinct descriptors $\le 2^{2^{|\cl(C_0)|}}$.
\item Kernel nodes: one per descriptor, at most $2^{2^{|\cl(C_0)|}}$.
\item Regular nodes: bounded by (blocking classes) $\times$ (tree
  depth).  The number of blocking classes is
  $|\Tp(C_0)| \times 4 \times 4^K$ where $K$ is the number of
  kernels.  With $K$ doubly exponential, this gives a tower of
  exponentials.
\item The V6 check (arc-consistency) runs in polynomial time in the
  network size.
\end{itemize}

The decision procedure is therefore computable but with a very high
complexity bound.  We do not claim \EXPTIME{} membership.

\begin{remark}[Towards tighter bounds]
The doubly-exponential bound on descriptors is likely a gross
overcount.  In practice, the number of distinct $\PP$-chain behaviors
is much smaller.  An \EXPTIME{} upper bound would require showing
that the quotient size is singly exponential, which we leave as an
open problem.  The lower bound is \EXPTIME-hard (inherited from
$\ALCI$).
\end{remark}


%% ============================================================
\section{Discussion}\label{sec:discussion}

\subsection{The role of full RCC5 tractability}

The soundness proof (Theorem~\ref{thm:soundness}) relies essentially
on Theorem~\ref{thm:full-tractability}: the full tractability of RCC5.
This deep result from qualitative spatial reasoning---that
path-consistency decides satisfiability for the entire RCC5
algebra---is what bridges the gap between the quotient's local
consistency (V5, V6) and the model's global consistency.

Without full tractability, we would need to verify global consistency
directly, which is the essence of the ``extension gap'' in earlier
approaches.  Full tractability collapses the gap: local consistency
(checkable) implies global consistency (needed).

\subsection{Why earlier approaches had gaps}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{p{3cm}p{3.5cm}p{5.5cm}}
\toprule
\textbf{Approach} & \textbf{Gap} & \textbf{Resolution in this paper} \\
\midrule
Quasimodel~\cite{Claude2026}
  & Extension gap: Henkin construction may fail
  & V6 + full tractability: path-consistent extensions always succeed \\
Contextual tableau~\cite{GPT2026}
  & FW$(C,N)$ false: infinite chains can't fit in fixed-width states
  & Period descriptors: chains as words, not graphs \\
Triangle types~\cite{ClaudeTriangle2026}
  & Extension solvability conjecture
  & V6 is extractable from models (model's assignment survives AC) \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Why the V6 argument works where Q3s failed}

The quasimodel approach required Q3s---arc-consistency on abstract
\emph{type-level} disjunctive networks.  Computational verification
showed that 11.1\% of model-derived type-level networks violate Q3s.
The key difference: V6 operates on \emph{instance-level} networks
with specific atomic relations between existing nodes, not on
abstract type-level domains.  The model provides actual relations for
each witness that satisfy \emph{all} composition constraints
simultaneously, so arc-consistency cannot eliminate them.

\subsection{Extension to $\ALCIRCC{8}$}

The argument relies on two properties of RCC5:
\begin{enumerate}
\item Full tractability (Theorem~\ref{thm:full-tractability}).
\item Universal self-absorption of $\PP$
  (Theorem~\ref{thm:self-absorbing}).
\end{enumerate}

For RCC8, $\PP$ splits into $\mathsf{TPP}$ (tangential) and
$\mathsf{NTPP}$ (non-tangential).  $\mathsf{NTPP}$ is transitive
($\comp(\mathsf{NTPP}, \mathsf{NTPP}) = \{\mathsf{NTPP}\}$), but
RCC8 is \emph{not} fully tractable---only specific subalgebras are
tractable~\cite{Renz1999}.  The decidability of $\ALCIRCC{8}$
remains open and may require additional techniques.


%% ============================================================
\section{Conclusion}\label{sec:conclusion}

We have proved that concept satisfiability in $\ALCIRCC{5}$ is
decidable.  The proof constructs a finite two-tier quotient for any
satisfiable concept: period descriptors capture the periodic structure
of infinite $\PP$-chains (Tier~1), and PP-kernel nodes with reflexive
$\PP$-loops capture cross-chain interactions (Tier~2).  Completeness
follows from the $T_\infty$ characterization of recurrent chain types
and the model's global consistency (which ensures the extension
condition V6).  Soundness follows from V6 and the full tractability
of RCC5 (path-consistent networks are satisfiable).

The decidability of $\ALCIRCC{5}$ has been an open problem since
Wessel~\cite{Wessel2003}.  Our resolution relies on two key
ingredients: (1)~the algebraic self-absorption of $\PP$ in the RCC5
composition table, and (2)~the patchwork property of RCC5 from
qualitative spatial reasoning.

The exact complexity remains open.  The decision procedure's upper
bound is computable but not elementary.  An \EXPTIME{} upper bound
(matching the $\ALCI$ lower bound) would require tighter analysis of
the quotient size.

The decidability of $\ALCIRCC{8}$ remains open, as RCC8 lacks full
tractability.


\begin{thebibliography}{9}

\bibitem{Wessel2003}
Michael Wessel.
\newblock \emph{Qualitative Spatial Reasoning with the ALCIRCC Family
  -- First Results and Unanswered Questions}.
\newblock Technical Report FBI-HH-M-324/03, University of Hamburg,
  2002/2003.

\bibitem{Claude2026}
Claude (Anthropic).
\newblock \emph{On the Decidability of ALCIRCC5 and ALCIRCC8:
  Quasimodels Meet the Patchwork Property}.
\newblock Discussion draft, March 2026.

\bibitem{GPT2026}
GPT-5.4 Pro.
\newblock \emph{A Contextual Tableau Calculus for
  $\mathcal{ALCI}_{\mathit{RCC5}}$}.
\newblock Discussion draft, 2026.

\bibitem{ClaudeFW2026}
Claude (Anthropic).
\newblock \emph{On the Failure of Finite-Width Extraction for
  $\mathcal{ALCI}_{\mathit{RCC5}}$}.
\newblock Discussion note, March 2026.

\bibitem{ClaudeTriangle2026}
Claude (Anthropic).
\newblock \emph{Triangle Types and the Extension Gap in
  $\mathcal{ALCI}_{\mathit{RCC5}}$}.
\newblock Discussion draft, March 2026.

\bibitem{RenzNebel1999}
Jochen Renz and Bernhard Nebel.
\newblock On the complexity of qualitative spatial reasoning:
  A~maximal tractable fragment of the Region Connection Calculus.
\newblock \emph{Artificial Intelligence}, 108(1--2):69--123, 1999.

\bibitem{Renz1999}
Jochen Renz.
\newblock Maximal tractable fragments of the Region Connection
  Calculus: A~complete analysis.
\newblock In \emph{Proc.\ IJCAI}, pages 448--454, 1999.

\bibitem{RandellCuiCohn1992}
David~A. Randell, Zhan Cui, and Anthony~G. Cohn.
\newblock A spatial logic based on regions and connection.
\newblock In \emph{Proc.\ KR}, pages 165--176, 1992.

\end{thebibliography}

\end{document}
